Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/heap.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, has children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\n{Creation, child comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, child promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, has
children}', `\n{Destruction, child comparison}' and `\n{Desruction child
promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\n{fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

@@

The most interesting thing is the relation between the predictability of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\n{Destruction, child comparison}' and `\n{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
the columns marked `\n{comparison 1}' to `\n{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsorts-3} and \ref{base and memory-tuned heapsorts-4} (the
8-heap).

@@

\begin{figure}
\includegraphics{plots/curve.eps}
\caption{Predictability of branches in memory-tuned heapsort, with a fanout of eight}
\label{memory-tuned heapsort curve}
\end{figure}

>>>>

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\n{Creation, Child Comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, Has
Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction Child
Promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\n{Fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

@@

The most interesting thing is the relation between the predictability of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
Promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
the columns marked `\n{Comparison 0}' to `\n{Child Promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsorts-3} and \ref{base and memory-tuned heapsorts-4} (the
8-heap).

@@

\begin{figure}
\includegraphics{plots/curve.eps}
\caption{Predictability of branches in memory-tuned heapsort, using an 8-heap.}
\label{memory-tuned heapsort curve}
\end{figure}

### 2005-08-14 docs/tech_report/merge.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{
Results shown are for a Direct Mapped cache.}{}{}

@@

Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. Base
mergesort has the same number of mispredictions (see Figure \ref{mergesorts-3},
indicating a possible minimum number. It has a lower number of branches,
probably due to performing a presort first. The presort has approximately 6.8
misses per key (not shown in the graphs). In later mergesorts, an insertion sort
is used instead.

Double-aligned tiled mergesort (which has the same results are tiled mergesort) has a
predilection for merging forwards. This is likely to be due to the fact that the
final merge goes forwards. The left merge step is also favoured, but this is
merely due to the fact that it occurs first in the code. While there are more
left branches, the actual number of branches taken are the same.

@@

Double multi-mergesort sorts the arrays fully in the cache, leaving them
bitonically arranged. This means that there is no bias for the forward branches,
as there was in tiled mergesort. This is shown in Figures \ref{mergesorts-5} and
\ref{mergesorts-6}. While merging the number of branches is 20\% lower than base
mergesort, while the number of mispredictions in only 10\% lower. This implies
that most of the mispredictions occur early in the merging process, and not in
the final merge, which multi-mergesort leaves out.

>>>>

\plot{merge}{}{}{Results shown are for a direct mapped cache.}{Results shown are
for a direct mapped cache.}{}{}

@@

Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. The
labels in the graphs correspond to the names of the steps in Knuth's algorithm,
which also correspond with the labels in our code.

Base mergesort has the same number of mispredictions as Knuth's sorts (see
Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
`\n{Right}' refer to interating through the first (or left-hand) and second (or
right-hand) lists being compared, respectively.

Base mergesort has a lower number of branches than Knuth's sorts, probably due
to performing a presort first. The presort has approximately 6.8 misses per key
(not shown in the graphs). In later mergesorts, an insertion sort is used
instead.

Double-aligned tiled mergesort (which has the same results are tiled mergesort)
has a predilection for merging forwards, shown in \ref{mergesorts-4}. The
`\n{Left}' and `\n{Right}' in the labels refer to the same as in
\ref{mergesorts-3}, while `\n{Forwards}' and `\n{Reverse}' refer to whether the
first or second part of the bitonic array segment is being merged.

This predilication is likely to be due to the fact that the final merge goes
forwards. The left merge step is also favoured, but this is merely due to the
fact that it occurs first in the code. While there are more left branches, the
actual number of branches taken are the same.

@@

Double multi-mergesort sorts the arrays fully in the cache, leaving them
bitonically arranged. This means that there is no final merge, and hence no bias
for the forward branches, as there was in tiled mergesort. This is shown in
Figures \ref{mergesorts-5} and \ref{mergesorts-6}. The labels in these graphs
are the same as in Figure \ref{mergesorts-4} and \ref{base and memory-tuned
heapsorts-2}.

While merging, the number of branches is 20\% lower than base mergesort, while
the number of mispredictions in only 10\% lower. This implies that most of the
mispredictions occur early in the merging process, and not in the final merge,
which multi-mergesort leaves out.

### 2005-08-14 docs/tech_report/quick.tex
fixed the captions for the graphs, and a few of the graphs themselves
added info about all the labels. Also capitalized them all

\plot{quick}{}{}{Results shown are for a Direct Mapped cache.}{}{}{}{}

@@

Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
the indiviual branches in base quicksort. The \cc{i} branch is the loop control
branch for the index sweeping from left-to-right up the array during partition.
The \cc{j} branch is the corresponding loop sweeping down the array. The
partition step terminates when $i \geq j$, and the \cc{partition end} branch
checks that condition. The \cc{insertion} branch is the comparison branch in the
insertion sort algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a final pass of
insertion sort over the whole array to sort these small partitions. Finally, the
\cc{median} branch is the sum of the results for the branches involved in
computing the median-of-three for each partition.

The only difference between cache quicksort and base quicksort is that
cache quicksort applies insertion sort to small partitions
immediately, whereas base quicksort does a single insertion sort pass
at the end. The result is that both algorithms perform almost
identically from the viewpoint of branch prediction. Cache quicksort
uses a very slightly smaller number of insertion sort comparison
branches, because it sorts only within partitions and does not attempt
to move array elements that have acted as pivots. (The reason that
quicksort works is that after partitioning, the pivot is in its
correct location, and must never move again). In practice, this
reduction in branches is probably offset by the increase in
mispredictions in the outer loop exit branch of the many insertion
sorts.

Base quicksort has a 35\% branch misprediction rate, whereas the rate
for cache quicksort is only 34\%. The main reason for this is the
reduction in branches by not attempting to move the pivot elements.
Attempting to move these elements usually results in a branch
misprediction. The main reason is that most other elements move at
least one position, and thus cause the insertion sort loop to iterate
at least once. But the pivot elements never move --- they are already
in the correct place.

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the \cc{i} and \cc{j} loops, with a corresponding reduction in the
number of mispredictions. Interestingly, the percentage misprediction rate of
the \cc{i} and \cc{j} branches does not change --- it is 37\% in both cases.

@@

Multi-quicksort does, however, introduce a lot of additional branches. We
already saw in Figure \ref{quick instructions}, that the binary search and
sequential search variants increase the number of executed instructions by
around 25\% and 50\% respectively. Figure \ref{base, memory-tuned and multi-
quicksorts-3} shows additional measures of the \cc{binary left} and \cc{binary
right} branches, which are the branches within the binary search code. The
number of binary search branches is much greater than the reduction in \cc{i}
and \cc{j} branches. Furthermore, these binary search branches are highly (43\%)
unpredictable.

The variation of multi-quicksort that uses sequential search dramatically
increases the number of executed branches, but these branches are highly
predictable. This version actually causes fewer branch mispredictions
than cache quicksort.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the \cc{i} and \cc{j}
branches in quicksort would resolve in each direction. In fact, this is not the
case. The main reason is that the chosen pivot is almost never the true median
of the section of the array to be partitioned. Thus, the \cc{i} and \cc{j}
branches are almost always biased, which makes them predictable. According to
Knuth, if the pivot element is chosen randomly, then the \cc{i} and \cc{j}
branches will be taken (i.e.  the \cc{i} or \cc{j} loop will execute) twice as
often as they are not taken (i.e. the loop exits).

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \cc{i} and \cc{j} branches are indeed taken about twice as often as
not-taken, and the \cc{i} and \cc{j} branches are correctly predicted around
71\% of the time.

The result that the branches are more predictable (71\%) than they are biased
(66\%) is a particularly interesting finding. The reason is that the bias in the
results is an average figure over a large number of partitions that quicksort
performs to sort the array. On a particular partition, the bias will be more
pronounced because the \cc{i} and \cc{j} branches will be biased in opposite
directions.

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \cc{i} loop will
execute a large number of times, and thus the \cc{i} loop will be biased towards
being taken. The \cc{j} loop, on the other hand, will be biased toward
not-taken, because the average element is likely to be smaller than the pivot.
Therefore, the \cc{j} loop will execute only a small number of times. Thus, we
see that in a particular partition, we do not normally get an equal number of
executions of the \cc{i} and \cc{j} loops. Instead we get a large number of
executions in one loop (which makes its branch relatively predictable), whereas
the other loop often does not enter at all (which makes its branch also
predictable, although executed a much smaller number of times). Thus, although
overall the \cc{i} and \cc{j} branches are taken twice as often as not-taken, on
a particular partition one will usually be biased in the not-taken direction,
and as a result will be executed a relativelty small number of times.

In the preceding paragraphs we have seen that the \cc{i} and \cc{j} branches are
predictable when we have a pivot that is larger or smaller than the true median
of the array section to be partitioned. An interesting question is the effect of
choosing a better approximation of the true median on branch prediction. Figure
\ref{base quicksort without a median, with median of 3, and with pseudo-medians
of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
used to select the pivot. Perhaps the biggest effect is that the number of
executions of the \cc{i} and \cc{j} branches falls by around 15\%. This fall is
partly offset by the comparison branches to compute the median, but these
account for only around 1.25\% of the original branches from the no-median
quicksort. Thus, there is a reduction of almost 14\% in the number of executed
comparison branches. Even more interesting, however, is the bias in the
remaining \cc{i} and \cc{j} branches. The no-median \cc{i} and \cc{j} branches
are 66\% biased, but the median-of-three branches are only 60\% biased in the
taken direction. The accuracy with which these branches are correctly predicted
falls from 71\% to 64\%.

>>>>

\plot{quick}{}{}{Results shown are for a direct mapped cache.}{Results shown are
for a direct mapped cache.}{}{}{}

@@

Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
the indiviual branches in base quicksort. The \n{i} branch is the loop control
branch for the index sweeping from left-to-right up the array during partition.
The \n{j} branch is the corresponding loop sweeping down the array. The
partition step terminates when \n{i $\geq$ j}, and the \n{Partition End} branch
checks that condition. The \n{Insertion} branch is the comparison branch in the
insertion sort algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a final pass of
insertion sort over the whole array to sort these small partitions. Finally, the
\n{Median} branch is the sum of the results for the branches involved in
computing the median-of-three for each partition.

The only difference between memory-tuned quicksort and base quicksort is that
memory-tuned quicksort applies insertion sort to small partitions immediately,
whereas base quicksort does a single insertion sort pass at the end. The result
is that both algorithms perform almost identically from the viewpoint of branch
prediction. Memory-tuned quicksort uses a very slightly smaller number of
insertion sort comparison branches, because it sorts only within partitions and
does not attempt to move array elements that have acted as pivots. (The reason
that quicksort works is that after partitioning, the pivot is in its correct
location, and must never move again). In practice, this reduction in branches is
probably offset by the increase in mispredictions in the outer loop exit branch
of the many insertion sorts.

Base quicksort has a 35\% branch misprediction rate, whereas the rate for
memory-tuned quicksort is only 34\%. The main reason for this is the reduction
in branches by not attempting to move the pivot elements.  Attempting to move
these elements usually results in a branch misprediction. The main reason is
that most other elements move at least one position, and thus cause the
insertion sort loop to iterate at least once. But the pivot elements never move
-- they are already in the correct place.

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the \n{i} and \n{j} loops, with a corresponding reduction in the
number of mispredictions. Interestingly, the percentage misprediction rate of
the \n{i} and \n{j} branches does not change -- it is 37\% in both cases.

@@

Multi-quicksort does, however, introduce a lot of additional branches. We
already saw in Figure \ref{quick instructions}, that the binary search and
sequential search variants increase the number of executed instructions by
around 25\% and 50\% respectively. Figure \ref{base, memory-tuned and multi-
quicksorts-3} shows additional measures of the \n{Binary Left} and \n{Binary
Right} branches, which are the branches within the binary search code. The
number of binary search branches is much greater than the reduction in \n{i}
and \n{j} branches. Furthermore, these binary search branches are highly (43\%)
unpredictable.

The variation of multi-quicksort that uses sequential search dramatically
increases the number of executed branches, but these branches are highly
predictable. This version actually causes fewer branch mispredictions than
memory-tuned quicksort.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the \n{i} and \n{j}
branches in quicksort would resolve in each direction. In fact, this is not the
case. The main reason is that the chosen pivot is almost never the true median
of the section of the array to be partitioned. Thus, the \n{i} and \n{j}
branches are almost always biased, which makes them predictable. According to
Knuth, if the pivot element is chosen randomly, then the \n{i} and \n{j}
branches will be taken (i.e.  the \n{i} or \n{j} loop will execute) twice as
often as they are not taken (i.e. the loop exits).

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \n{i} and \n{j} branches are indeed taken about twice as often as
not-taken, and the \n{i} and \n{j} branches are correctly predicted around 71\%
of the time.

The result that the branches are more predictable (71\%) than they are biased
(66\%) is a particularly interesting finding. The reason is that the bias in the
results is an average figure over a large number of partitions that quicksort
performs to sort the array. On a particular partition, the bias will be more
pronounced because the \n{i} and \n{j} branches will be biased in opposite
directions.

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \n{i} loop will
execute a large number of times, and thus the \n{i} loop will be biased towards
being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
because the average element is likely to be smaller than the pivot.  Therefore,
the \n{j} loop will execute only a small number of times. Thus, we see that in a
particular partition, we do not normally get an equal number of executions of
the \n{i} and \n{j} loops. Instead we get a large number of executions in one
loop (which makes its branch relatively predictable), whereas the other loop
often does not enter at all (which makes its branch also predictable, although
executed a much smaller number of times). Thus, although overall the \n{i} and
\n{j} branches are taken twice as often as not-taken, on a particular partition
one will usually be biased in the not-taken direction, and as a result will be
executed a relativelty small number of times.

In the preceding paragraphs we have seen that the \n{i} and \n{j} branches are
predictable when we have a pivot that is larger or smaller than the true median
of the array section to be partitioned. An interesting question is the effect of
choosing a better approximation of the true median on branch prediction. Figure
\ref{base quicksort without a median, with median of 3, and with pseudo-medians
of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
used to select the pivot. Perhaps the biggest effect is that the number of
executions of the \n{i} and \n{j} branches falls by around 15\%. This fall is
partly offset by the comparison branches to compute the median, but these
account for only around 1.25\% of the original branches from the no-median
quicksort. Thus, there is a reduction of almost 14\% in the number of executed
comparison branches. Even more interesting, however, is the bias in the
remaining \n{i} and \n{j} branches. The no-median \n{i} and \n{j} branches are
66\% biased, but the median-of-three branches are only 60\% biased in the taken
direction. The accuracy with which these branches are correctly predicted falls
from 71\% to 64\%.

### 2005-08-14 docs/tech_report/shell.tex
fixed captitalization in captions
added info about all the labels. Also capitalized them all

Two versions of Gonnet's sort were used. The first was taken directly from the
code presented by Sedgewick, and only slightly modified to fit into the
framework used.  The second was modified to reduce the complexity of the loop
bounds, and to remove a sentinel check in the inner loop. Code for shellsort and
improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
Shellsort code}.

\code{Shellsort code}{shellsort.c}
\code{Improved Shellsort code}{improved_shellsort.c}

@@

The results also show that in shellsort, a key moves only 1.03 places down the
array, on average, during each iteration. In improved shellsort, each key moves
over 1.07\% places each iteration, which in the final insertion sort, each key
moves an average of 2.1 places.

>>>>

Two versions of Gonnet's sort were used. The first was taken directly from the
code presented by Sedgewick, and only slightly modified to fit into the
framework used.  The second was modified to reduce the complexity of the loop
bounds, and to remove a sentinel check in the inner loop. Code for shellsort and
improved shellsort is shown in Figures \ref{Shellsort code} and \ref{Improved
shellsort code}.

\code{Shellsort code}{shellsort.c}
\code{Improved shellsort code}{improved_shellsort.c}

@@

The results also show that in shellsort, a key moves only 1.03 places down the
array, on average, during each iteration. In improved shellsort, each key moves
over 1.07\% places each iteration, while in the final insertion sort, each key
moves an average of 2.1 places.
